---
title: "Lab_2_markdown"
author: "Zane Shango, Kurtis Potier, Nick Hiller"
date: "1/28/2020"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in Ames Data and Print Variable Names

```{r, out.width="500px", echo = TRUE}
ameslist <- as.data.frame(read.csv("https://msudataanalytics.github.io/SSC442/Labs/data/ames.csv", header =TRUE,sep = ","))

names(ameslist)
```
## Questions Leading up to Exercise 1
Interpretations for confusing variables are:
LotShape describes what the lot looks like, reg likely means rectangular and IR1, IR2, etc. are for irregular shapes
MSSubClass is the type of dwelling in the sale according to googling it.
The conditions do not provide enough information to say. It may be useful to check if they cause differnces in the sale price to see if they can be ignored.
ameslist is of type list.

```{r,  out.width="500px", echo = TRUE}

typeof(ameslist)
```

## GarageTemp and Update ameslist to only 0 and 1 Values outputs
```{r,  out.width="500px", echo = TRUE}

library(rowr)

GarageTemp <- model.matrix( ~ GarageType - 1, data=ameslist )
ameslist <- cbind.fill(ameslist, GarageTemp)

ameslist$GarageOutside <- ifelse(ameslist$GarageTypeDetchd == 1 | ameslist$GarageTypeCarPort == 1, 1, 0)
unique(ameslist$GarageOutside)
# there are no NA values, just 0s and 1s

```

## Exercise 1
```{r,  out.width="500px", echo = TRUE}

Ames<-ameslist[which(sapply(ameslist,is.numeric)==TRUE)]

names(Ames)
pairs(~LotArea+YearBuilt+OverallQual+OverallCond+BsmtFinSF1+TotRmsAbvGrd+MoSold+FullBath+
        HalfBath+GarageCars+BedroomAbvGr+YrSold,data=Ames)

Ames_small <- Ames[,c(4,5,6,7,10,20,21,22,24,27,36,37,38)]


head(Ames_small)

cor(Ames_small)


```
Most matched prior beliefs, overall condition and sale price were negatively correlated, which was surprising

```{r,  out.width="500px", echo = TRUE}
library(ggplot2)

reg<-lm(SalePrice~GrLivArea,data=ameslist)
ggplot(ameslist,aes(x=GrLivArea,y=SalePrice))+
  geom_point()+
  geom_abline(intercept=coef(reg)["(Intercept)"],slope=coef(reg)["GrLivArea"])
reg

```

The following line gives information about the largest outlier above the line
```{r,  out.width="500px", echo = TRUE}
ameslist[which.max(ameslist$SalePrice-(ameslist$GrLivArea*coef(reg)["GrLivArea"]+coef(reg)["(Intercept)"])),]

```

## Exercise 2

Linear Model
```{r,  out.width="500px", echo = TRUE}
attach(Ames)
lm.fit <- lm(SalePrice ~ GrLivArea)
lm.fit
```

If GrLivArea is Omitted? Check Correlation
```{r,  out.width="500px", echo = TRUE}
cor(ameslist$GrLivArea,Ames_small)
lm.fit
summary(lm.fit)
```

Plot: there are some outliers but the data seems to be fairly normally grouped
```{r,  out.width="500px", echo = TRUE}
plot(lm.fit)

```

Add LotArea to Model
```{r,  out.width="500px", echo = TRUE}
lm.fit = lm(SalePrice ~ GrLivArea + LotArea)
lm.fit
```
# Does controlling for LotArea change the qualitative conclusions from the previous regression?
Not very much, a greater ground living area still means greater sale price and is still very significant.

# What about the quantitative results? 
The coefficient went from 107 to 104 on ground living area.

# Does the direction of the change in the quantitative results make sense to you?
Yes, adding lot area will reduce the coefficient since lotarea and grlivarea are related.


## Value of Indoor Garage
An indoor garage is worth $2849, however the p value does not indicate significance.
```{r,  out.width="500px", echo = TRUE}
summary(lm(SalePrice ~ GarageOutside,data=ameslist))

```

## Multiple Linear Regression
```{r,  out.width="500px", echo = TRUE}
all<-lm(SalePrice~.,data=Ames)
summary(all)

```

# Is there a relationship between the predictors and the response? 
# Which predictors appear to have a statistically significant relationship to the response?
Yes, there is a relationship.
MSSubClass,LotArea, OverallQual, OverallCond, YearBuilt,MasVnrArea,BsmtFinSF1, X1stFlrSF, X2ndFlrSF,
BsmtFullBath, BedroomAbvGr, KitchenAbvGr, TotRmsAbvGrd, Fireplaces, GarageCars, and ScreenPorch appear to have a statistically significant relationship.

# What does the coefficient for the year variable suggest?
The trend of housing prices was downward for the duration of the data collection.

## Diagnostic Plots of Linear Regression Fit
```{r,  out.width="500px", echo = TRUE}
plot(all)

```

# Problems with the Fit? Any Unusually High Leverages?
There are several large outliers;one observation has very high leverage, some others have moderately high leverage.

# Recall that the operator : designates the interaction between two variables. The operator * designates the interaction between the two variables, plus the main effects.

## Find Statistically Significant Interactions
```{r,  out.width="500px", echo = TRUE}
inter1<-lm(SalePrice~GrLivArea*FullBath,data=ameslist)
summary(inter1)

inter2<-lm(SalePrice~LotFrontage:LotArea,data=ameslist)
summary(inter2)

```

## Transformations of the Variables
# Baseline
```{r,  out.width="500px", echo = TRUE}
baseline <- lm(SalePrice ~ LotArea,data=ameslist)
summary(baseline)

```
# Ln Transformation
The t value improved from 10 to 16 when using an ln function  and the r^2 also got much better, going from 0.07 to 0.15.
```{r,  out.width="500px", echo = TRUE}
trans_ln <- lm(SalePrice ~ log(LotArea),data=ameslist)
summary(trans_ln)
summary(baseline)

```
# Squared Transformation
There is no difference in result when the value is squared
```{r,  out.width="500px", echo = TRUE}
trans_x2 <- lm(SalePrice ~ (LotArea*LotArea),data=ameslist)
summary(trans_x2)
summary(baseline)

```
# Sqrt Transformation
Using the squareroot, the r^2 value got better, going from .07 to .14 and the t value also improved from 10 to 15.
```{r,  out.width="500px", echo = TRUE}
trans_sqrt <- lm(SalePrice ~ sqrt(LotArea),data=ameslist)
summary(trans_sqrt)
summary(baseline)
```

